The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-06-30 Paper/jaccardbaetica.tex
comentarios y edits

The aim of this study is to explore economics dynamics in the production and distribution of olive oil trade. 
Our case of study has been focused on the production processes located in \textit{Baetica} province (currently Andalusia) from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. \textit{Baetica} became an important production and distribution centre during the Roman Empire. However, it remains under debate about how this province was organised and whether it could be possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from two centres: 1) production centres by analysing different workshops in \textit{Baetica} province and 2) two Roman provinces such as \textit{Germania} and \textit{Britannia} as consumption centres. They will be used to detect a connection between the distribution of amphoric stamps and the economic structure in both centres. Here, we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

@@

The role of the Mediterranean sea for the trade had an enormous impact during the Roman Empire. The spread of thousands of different goods along maritime shipping allowed progressive exploitation of communication networks and a major interaction between communities \citep{rodriguez_baetican_1998, 
temin_market_2001,
bevan_mediterranean_2014}. The exchange of different products between regions can be seen in both archaeological and written records as a frequent indicator of Roman trade.

@@

Given this fact, an important mechanism of production control under the Roman government was created spreading by different areas attending to the richness of different places. Roman provinces such as \textit{Baetica} (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered the liquid gold since it was used in different aspect of daily life for cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence of high demand, this product was distributed in a large number of amphorae across the provinces, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

The understanding of the olive oil production processes has been widely discussed over many decades\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. Despite this topic is still subject to debate: what patterns were followed for the distribution of olive oil to the different provinces?
Did each province follow a different pattern for its distribution? Nor the lack of written records has made possible to detect any indication of patterns in the olive oil market.

Advances in the research of the Roman studies have currently led an environment with more diverse commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}. The application of different quantitative approaches combined with the archaeological evidence has allowed us partly to deal with the complexity of the Roman production \citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.

This paper aims to study the olive oil market connection between provinces by calculating the similarity of stamps. Specifically, our work pretends to detect microeconomic processes focused on a commercial product from a specific province \citep{isaksen_network_2006}. We want to understand the pattern of olive oil production linked to amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production and distribution centres.

@@

\section{The Amphoric production in Baetica province}

The high demand of Roman provinces stimulated by the good condition of the \textit{Baetica} lands allowed to develop a massive infrastructure of olive oil production. The production and distribution of olive oil were growing exponentially for almost three centuries \citep{remesal_concierto}. As a result, hundreds of amphora workshops were raised to support this high demand of the Roman Empire. The amphora workshops were situated in different locations in \textit{Baetica} province, along the rivers Guadalquivir and Genil (see Fig.\ref{workshop}).

The strategic location of the workshops at the riverbank allowed the transport of olive oil through riverine transport towards the maritime routes connecting both the Mediterranean and the Atlantic \citep{garcia_vargas_enrique_formal_2010}.

@@

The chronology in the workshops is widely diverse from the first to the third centuries AD \citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. 
This fact is shown by the archaeological evidence that displays a highly specialised production with a long activity with apparently few changes \citep{remesal_anforas_2004}.

The majority of olive oil amphorae produced in this province belong to Dressel 20 type\citep{dressel_ricerche_1878,
millet_anforas_1998}. Dressel 20 amphorae were shipped and distributed from \textit{Baetica} throughout western Europe.

Dressel 20 contains much information still unknown. A high percentage of  Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in \textit{graffiti}. Indeed, most of the inscriptions on the amphora have not enough analysed due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}.

@@

Dressel 20 was the amphora most stamped during the Roman Empire \citep[18]{millet_anforas_1998}. The usage period is associated with a large amphora production for almost three centuries.

Most stamps display a large activity of production that it can be difficult for specifying an accurate chronology in some cases. They have been frequently dated with the consular dating by studying \textit{tituli picti} found in \textit{Monte Testaccio} \citep{Testaccio1, berni_millet_epigrafianforica_2008}.
However, the chronology of stamps could be also biased in other sites taking into account that amphorae were deposited after being produced and consumed.

The fascination aroused by stamps has caused it to be one of the inscriptions with the most publications about its origin and meaning. Its endure in time become an essential element that was used to mark amphorae during almost three centuries \citep{remesal_sellar_2016} (Fig.\ref{amphora}).

@@

Frequently, they were mainly marked in handles, but seldom in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion. Stamps are mostly formed by a code of three letters and they can appear in abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

Despite researchers identify stamps as identity marks, there is not a consensus about the meaning of the stamps \citep{rodriguez_baetican_1998}. Alike a high percentage of Dressel 20 containers were not stamped being difficult to determinate.

They could be identified based on three premises: content (olive oil), context (amphora workshop) and subject (individuals involved). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other, they could belong to the owner of the making amphorae workshops or even a group of amphora workers \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps became in a good proxy to define somehow the system of working in the workshops.

Nevertheless, some challenges remain under discussion such as how this production was organised and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop?

Neither the use of written records have allowed providing enough information that can explain the economic role of \textit{Baetica} province in the Roman organisation.

@@

The conquest of new provinces allowed the Roman Empire the arrival of resources through Mediterranean and Atlantic routes. This lead to a progressive change in the economic and social structure creating new trade networks for supplying Roman conquest. As a result, Augusto's administration created the figure of the \textit{praefectura annonae} for the supply of wheat. The role of the \textit{praefectura annonae} was mainly to provide though \textit{frumentationes} a quantity per month of wheat to Roman citizens \citep{remesal_annona_1986,remesal_concierto}.

\textit{Annona} could have also included other goods for the supply of Roman legions such as olive oil according for some hypotheses \citep{remesal_annona_1986,remesal_annona_1990}

The importance of olive oil production could be explained by the fact that we detected a massive amphora distribution in different Roman provinces. This amphora has been commonly associated with the  transportation of Baetican olive oil for supplying military camps and civil settlements \citep{berni_millet_epigrafianforica_2008}.

The military consumption of olive oil in Roman provinces may be related to two aspects: a) cultural consumption whose product is consumed by cultural reasons such as to identity or habit and b) economical reasons where olive oil distribution is based on the transport costs, unlike other products \citep[69-70]{carreras_britannia_1998}.

Olive oil supply had a special impact in militarised provinces such as \textit{Britannia} and \textit{Germania}. The concentration of amphorae in eminently militarised areas suggests the existence of a broad supply mechanism, being accurate to imagine a possible organisation in charge of its distribution from military camps to civil settlements\citep{remesal_annona_1986, carreras_britannia_1998}.

@@

The consumption of olive oil in \textit{Britannia} was residual until starting the Roman conquest \citep{funari_corpus_1996,
carreras_abastecimiento_2003}. It is well known that this product was not frequently consumed by indigenous population \citep[161]{monfort_britanniaen_1998}. The absence of olive oil importation before the conquest can be reflected by the shortage of this product until the arrival of a high percentage of Dressel 20 amphorae to \textit{Britannia} \citep[1]{carreras_britannia_1998}. Additionally, the land of \textit{Britannia} showed inadequate for the olive oil production due to low environmental conditions. This issue was necessary for creating an important export apparatus from \textit{Baetica} to supply Roman army.

At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}. This fact will have a spatial incidence in sites close at Hadrian Wall.  
Olive oil production in \textit{Baetica} would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic points \citep{carreras_atlantic_2012}. The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during military campaigns.

The presence of Dressel 20 stamps in military camps in \textit{Britannia} have been widely studied in Roman archaeology \citep{carreras_britannia_1998}. This fact also indicates a possible governmental organisation allocate to reinforce and supply of olive oil in military camps, such as \textit{Germania} (Remesal Rodrıguez, 1986). However, it is unknown the type of economical system when managing or control the redistribution of olive oil in different sites in \textit{Britannia} but it seems that it could have been in the main cities \citep[45]{funari_economic_2005}.

Finally, the increase in olive oil exports would experience a progressive slowdown from the third century A.D., coinciding with the change in market strategy in the Empire. As a result, a progressive decrease of Dressel 20 is documented being gradually replaced for Dressel 23 amphorae \citep{rodriguez1991aceite,millet_anforas_1998}.

>>>>

\xavi{reescribir abstract, demasiado general y entonces demasiado específico}
The aim of this study is to explore economics dynamics in the production and distribution of olive oil trade. 
Our case of study has been focused on the production processes located in \textit{Baetica} province (currently Andalusia) from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. \textit{Baetica} became an important production and distribution centre during the Roman Empire. However, it remains under debate about how this province was organised and whether it could be possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from two centres: 1) production centres by analysing different workshops in \textit{Baetica} province and 2) two Roman provinces such as \textit{Germania} and \textit{Britannia} as consumption centres. They will be used to detect a connection between the distribution of amphoric stamps and the economic structure in both centres. Here, we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

@@

The intensification of long-range trade was one of the most important traits of the economy developed duriong the Roman Empire. The development of an extensive road network increased the connectivity between inland communities, but most shipping continued being based on maritime routes particularly in the Mediterranean basin~\citep{temin_market_2001,bevan_mediterranean_2014}. The intensity of this trade between distant regions can be observed both in archaeological and written evidence~\citep{rodriguez_baetican_1998}.

@@

The Empire developed a series of structures to support and organize this long-range trade and specialized entire regions to massively produce specific goods. An example of this process was the province of \textit{Baetica} (currently Andalusia, southern Spain) which became an important olive oil production centre during\xavi{período?}. Olive oil was an essential good for Romans because it was used in almost every aspect of their daily life such as cooking, hygiene or lighting~\citep{mattingly_d.j._oil_1988}. This high demand required a huge increase on the production which was distributed through amphorae shipped via maritime routes to all the provinces and particularly to Italy and the hundreds of garrisons that the Roman army deployed along its borders~\citep{blazquez_exportacion_1980}.

The structure and processes associated to this massive olive oil production has been extensively discussed over the last decates\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. Our knowledge of this economic activity has benefitted from new findings and data sources, but despite all these advances several questions remain open: \xavi{no entiendo esta pregunta} what patterns were followed for the distribution of olive oil to the different provinces? \xavi{no es esta pregunta idéntica a la anterior?} Did each province follow a different pattern for its distribution? \xavi{esta frase no tiene sentido aquí; o la pones antes o en otro párrafo, pero es raro acabar un set de 2 preguntas con una frase que no sea otra pregunta} Nor the lack of written records has made possible to detect any indication of patterns in the olive oil market.

\xavi{las preguntas no van sobre patrones; los patrones son las trazas en la evidencia que te ayudan a responder a las preguntas}

Advances in the research of the Roman studies have currently led an environment with more diverse interpretations on commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}\xavi{more diverse que qué?}. The application of different quantitative approaches has allowed us to improve our interpretation of the growing amount of archaeological evidence and reveal the complexity of the Roman production \xavi{en qué sentido? qué quieres decir aquí con complexity? Si no especificas queda muy genérico (rollo "el mundo es muy complejo")}.\citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.

\xavi{me parece que aquí falta más chicha, porque saltas de una cosa super general (complejidad, economía, grandes dinámicas) a unos casos muy específicos. Deberías aquí explicar algo más sobre tu estudio y lo que quieres entender, por ejemplo qué preguntas específicas quieres responder? Como se enlazan con lo que has explicado hasta ahora? Cómo es que ahora saltas a casos de estudio regionales? Las preguntas específicas de los 3 casos de estudio deberían entrelazarse en una pregunta general del paper, que esté vinculada a lo que has explicado hasta ahora}

This paper aims to study the olive oil market connection between provinces by calculating the similarity of stamps. Specifically, our work pretends to detect microeconomic processes\xavi{no me gusta lo de microeconomic, porque es un campo específico de la economía que nada tiene que ver con analizar un limes entero. Quizás regional?} focused on a commercial product from a specific province \citep{isaksen_network_2006}. We want to understand the pattern of olive oil production linked to amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production and distribution centres.

@@

\section{The Amphoric production of Baetica}

The landscape of the \textit{Baetican} province was one of the best regions to face the increase in demand of olive oil across the Roman provinces. For this reason the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deplyoyed. The production and distribution of olive oil grew exponentially\xavi{seguro que exponencial?} for almost three centuries\xavi{qué centuries?} \citep{remesal_concierto}. Hundreds of presses and amphora-making workshops were build near large extensions covered by olive trees. The workshops where amphorae were made and filled with olive oil were places along the rivers Guadalquivir and Genil riverbanks (see Fig.\ref{workshop}).

This strategic location allowed the transport of olive oil through riverine shipping towards the maritime routes that connected Baetica with Mediterranean and Atlantic trade routes towards the rest of the Empire\citep{garcia_vargas_enrique_formal_2010}.

@@

The chronology of the workshops is widely diverse from the first to the third centuries AD\xavi{qué quiere decir que la cronología es diversa?} \citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. 
This fact is shown by the archaeological evidence that displays a highly specialised production with a long activity with apparently few changes\xavi{deberías desarrollar esto un poco más; qué tipo de evidencia demuestra esto? pocos cambios en qué? los talleres? las tipologías anfóricas?} \citep{remesal_anforas_2004}.

A large majority of this Baetican amphorae production is composed by Dressel 20 types\citep{dressel_ricerche_1878,millet_anforas_1998}. Dressel 20 amphorae were shipped and distributed from \textit{Baetica} throughout western Europe\xavi{esto ya lo has mencionado; quizás podrías explicar en 2 líneas la capacidad y forma de la dressel 20}.

Dressel 20 amphorae display a lot of evidence, but some of it is still hard to interpret. A large percentage of Dressel 20 were marked with stamps, while they could also be inked with \textit{tituli picti} or incised with \textit{graffiti}. The interpretation for most of these inscriptions is still open due to the fragmentation of the material and the small sample size of well-preserved elements~\citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}.

@@

Dressel 20 was the amphora most stamped during the Roman Empire \citep[18]{millet_anforas_1998}. The usage period is associated with a large amphora production for almost three centuries\xavi{esto está repetido no?}.

The large sample size of recovered Dressel 20 stamps is distributed across all the western provinces of the Roman Empire. For this reason there are several publications of this type of evidence discussing its origin, its meaning, its long period of use and its spatial distribution~\citep{remesal_sellar_2016} (Fig.\ref{amphora}).\xavi{si dices que es muy popular debería haber aquí más citas aparte de Remesal}.

@@

Most stamp codes display a large activity of production that can be difficult to date using an accurate chronology. They have been frequently dated with the consular dating by studying \textit{tituli picti} found in \textit{Monte Testaccio} \citep{Testaccio1, berni_millet_epigrafianforica_2008}. However, the chronology of stamps could be also biased in other sites taking into account that amphorae were deposited after being produced and consumed\xavi{esto necesita un poco de desarrollo. quieres decir que el código se encuentra en contextos de cronologías diferentes no? Y lo del Testaccio, pos eso, que hay pocos y todos de un yacimioento, lo que añade un bias}.

These stamps are typically found on handles but they were also imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion\xavi{qué information? El código de letras?}. Stamps often displayed a code of three letters and they can appear in abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.\xavi{que quieres decir con abbreviated o complete? Como abrevias un código de 3 letras?}

Scholars agree that stamps are some type of identification mark, but there is no consensus on the meaning of these marks\citep{rodriguez_baetican_1998}. Additionally, a large percentage of Dressel 20 containers were not stamped thus complicating the interpretation~\xavi{esta última frase aquí es un poco random no?}. The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved on the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops were the amphorae was made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps is a good proxy to explore the system of olive oil production and trade developed in Baetica.

Nevertheless, some challenges\xavi{acabas de decir que tampoco se sabe qué es el sello, así que esta frase queda rara} remain under discussion such as how this production was organised and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop?

Neither the use of written records have allowed providing enough information that can explain the economic role of \textit{Baetica} province in the Roman organisation.\xavi{esta frase esta fuera de contexto. Si la pones entonces deberías expandir la idea de frase a párrafo, y poner referencias}.

@@

\xavi{este título es muy raro, porque 1. que son casos de estudio ya lo sabes y 2. case study singular, consumption centres plural}

The creation of new provinces allowed the Roman Empire the arrival of resources through Mediterranean and Atlantic routes\xavi{no entiendo esto...allowed o required?}. This led to a gradual change in the economic and social structureas the trade networks that supplied the Roman army became more complex. Augustus's administration created the figure of the \textit{praefectura annonae} to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}\xavi{pero estabas hablando del ejército no? Has hecho un salto aquí a los ciudadanos}.

Some authors argue elsewhere that this same \textit{Annona} could also have organized the supply of additional goods such as olive oil to the Roman legions \citep{remesal_annona_1986,remesal_annona_1990}\xavi{pero no hay evidencia? Las fuentes escritas no lo mencionan?}.

The importance of this olive oil trade is revealed by the massive amount of Dressel 20 amphorae found across the Roman provinces. This amphora has been commonly associated with the transportation of Baetican olive oil for supplying military camps and civil settlements \citep{berni_millet_epigrafianforica_2008}.\xavi{me parece repetitivo, porque unos párrafos antes ya hablabas de lo mismo. Quizás unificar texto aquí o allá?}

The military consumption of olive oil in Roman provinces may be related to two aspects: a) cultural consumption whose product is consumed by cultural reasons such as to identity or habit and b) economical reasons where olive oil distribution is based on the transport costs, unlike other products \citep[69-70]{carreras_britannia_1998}\xavi{no pillo esta frase}..

Olive oil supply seems to be particularly intense in militarised provinces that functioned as the borders of the Empire such as \textit{Britannia} and \textit{Germania}. The high percentage of amphorae\xavi{amphorae o Dressel 20?} found in sites nearby military garrisons suggests the existence of diverse supply dynamics as there were trade processes flowing good from military sites to civil settlements\citep{remesal_annona_1986, carreras_britannia_1998}\xavi{lo que quieres decir es que tienes muchas ánforas en sitios civiles cerca de militares? no queda claro}.

\xavi{faltaría un enlace a la siguiente sección, queda muy abrupto}

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. Indigenous\xavi{no sé si indigenous es la mejor palabra...local?} population did not consume this product and it was not produced on the region as the landscape and climate of the British isles was not suitable for olive oil trees\citep[161]{monfort_britanniaen_1998}. Based on the archaeological record this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} \citep[1]{carreras_britannia_1998}\xavi{en qué siglo?}. The demand of olive oil after the Roman conquest required of a new trade route due to the lack of local olive oil production; \textit{Baetica} was chosen as its main supply.

At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}\xavi{repetitivo? fusionar párrafos}. This fact will have a particular intensity within sites close to the Hadrian Wall's garrisons.  Olive oil production in \textit{Baetica} would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic locations~\citep{carreras_atlantic_2012}. The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during military campaigns\xavi{estas 2 frases repiten cosas ya explicadas}.

The presence of Dressel 20 stamps in military camps in \textit{Britannia} has been widely studied in Roman archaeology \citep{carreras_britannia_1998}.\xavi{si dices widely studied no puedes referencias solo 1 autor}. This fact\xavi{que fact? Que los sellos de dressel 20 han sido muy estudiados? No tengas miedo a repetir trozos rollo "This intense consumption of olive oil also indicatas blablabla...} also indicates a possible provincial structure designed to organize the supply of olive oil to military camps,m such as \textit{Germania} (Remesal Rodrıguez, 1986)\xavi{por qué such as Germania? Aún no has explicado qué hacen en Germania}. There are no written records explaining how this redistribution of essential goods was organised in \textit{Britannia}, but the archaeological evidence suggests that cities may have been the central nodes of this local trade network~\citep[45]{funari_economic_2005}.

The consumption of olive oil would experience a progressive slowdown from the third century A.D. onwards. This date matches a change in market strategy in the Empire. This a gradual decrease on \textit{Baetican} olive oil exports can be observed as a decrease in the amount of Dressel 20 found in excavated sites as this typology is replaced by Dressel 23 amphorae \citep{rodriguez1991aceite,millet_anforas_1998}.\xavi{pero entonces la dressel 23 de donde viene? Cuidao que estás diciendo que se acaban las dressel 20, pero esto implica que hay menos aceite de oliva o que viene de otras partes o que va en recipientes distintos a la dressel 20?}

\xavi{HE HECHO HASTA AQUÍ}
